---
title: "Getting started with bayesrep"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting started with bayesrep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5,
    fig.align = "center"
)
options(width = 90)
library(bayesrep)
```

## 1. What this package is for

You have many predictors, possibly more than observations, and many of them are
correlated with each other.

A standard sparse model (lasso, spike-and-slab) handles this by **picking one
predictor out of each correlated group and throwing the rest away**. The choice
is close to arbitrary: resample your data and you often get a different winner.
Coefficients are unstable and uncertainty is understated.

`bayesrep` fits an **ensemble of `G` sparse sub-models at once**. Two things
act together:

* every sub-model has to predict the response **on its own**, so each needs a
  competent set of predictors;
* a **repulsive prior** penalises a sub-model for reaching for a predictor
  another sub-model already holds.

On a correlated block the result is that the block gets *split across*
sub-models, each reconstructing the signal through its own member. Predictions
average the sub-models. You get full Bayesian uncertainty, and an explicit,
inspectable account of which predictors are acting as substitutes for one
another.

## 2. A worked example

Eight predictors. `V1` and `V2` are nearly identical (correlation above 0.98)
and both genuinely matter. `V3` matters and is on its own. The rest are noise.

```{r data}
set.seed(42)
n <- 120
p <- 8

X <- matrix(rnorm(n * p), n, p)
X[, 2] <- X[, 1] + rnorm(n, sd = 0.15)      # V2 is almost a copy of V1
colnames(X) <- paste0("V", 1:p)

# truth: intercept 10, V1 = 2, V2 = 2, V3 = -3, V4..V8 = 0
y <- as.numeric(10 + X %*% c(2, 2, -3, rep(0, 5)) + rnorm(n))

round(cor(X[, 1], X[, 2]), 3)
```

Fitting is one call.

```{r fit}
fit <- bre(X, y, numModels = 3, iter = 3000, burnin = 1500,
           verbose = FALSE, seed = 7)
```

The two settings worth thinking about are `numModels` (how many sub-models) and
`lambda2` (how hard they repel). The default `lambda2` is deliberately
conservative and on this small example it is **too weak to separate anything** —
every feature ends up in every sub-model:

```{r default-weak}
rowSums(fit$pip)[1:3]        # average number of sub-models carrying V1, V2, V3
```

Sections 3 to 5 use this fit to show the mechanics, which are unaffected.
Section 6 then chooses `lambda2` properly, and Section 7 uses that fit to look
at structure. **Do not skip Section 6 when you apply this to your own data.**

## 3. The three things to look at

In order: **did it run properly**, **what did it find**, **should I believe it**.

### `print(fit)` — did it run properly?

```{r print}
fit
```

What to check:

| Line | What you want |
|---|---|
| `Transfer acc.` | comfortably above zero. This is the move that lets features change sub-model; if it is 0% the assignment is frozen. |
| `theta MH acc.` | high is fine and expected. |
| `Sub-model size` | not all zeros (nothing found) and not everything (no sparsity). |
| `Selected (FDR 10%)` | a plausible number of real signals. |

### `summary(fit)` — what did it find?

This is the main output of the package.

```{r summary}
summary(fit)
```

Column by column:

* **`eMIP`** — "ensemble marginal inclusion probability": the posterior
  probability that the feature is used by **at least one** sub-model. This is
  your variable-importance score, and a correct per-feature probability.
  Use it to **rank**, not to threshold -- see the warning below.
* **`PostMean`, `Lower`, `Upper`** — the ensemble coefficient and its 95%
  credible interval, **in your original units**. No back-transforming needed.
* **`Theta`** — how broadly useful the feature is, on a 0–1 scale.
* **`MeanNModels`** — the average number of sub-models carrying the feature.
  Near 1 means it lives in a single sub-model; near `G` means every sub-model
  wants it (a "super-predictor" the repulsion could not push apart).

Notice `V1` and `V2`. Both have high `eMIP`, so the model is certain they
matter, but their individual intervals are wide because the data genuinely
cannot separate them. A single sparse model would typically have kept one and
zeroed the other, hiding that ambiguity. Only their **sum** is well identified:

```{r summary-pair}
cf <- coef(fit)
cat("V1 + V2 estimated:", round(sum(cf[c("V1", "V2")]), 3), " true: 4\n")
cat("V3 estimated:", round(cf[["V3"]], 3), " true: -3\n")
```

### Selecting features: do not threshold eMIP

It is tempting to call everything with `eMIP > 0.5` a discovery. **Don't.**
eMIP is the probability that a feature is carried by *at least one* of the `G`
sub-models, so a null feature gets `G` chances to be picked up. A feature whose
per-sub-model inclusion probability is only 0.12 has
`1 - (1 - 0.12)^5 = 0.47` at `G = 5` — it crosses 0.5 on arithmetic alone. In a
simulation with 350 null predictors, `eMIP > 0.5` admitted **347 of them**.

Thresholding the per-sub-model probabilities instead does not help: the
posterior is invariant to permuting sub-model labels, so a feature reliably
carried by exactly one of five sub-models sits near `1/5` in each and never
reaches 0.5 anywhere.

Use `selectFeatures()`, which ranks by eMIP and admits features while the
estimated false discovery rate — the average of `1 - eMIP` over those selected
— stays within your target. That quantity never refers to `G`:

```{r select}
selectFeatures(fit, fdr = 0.1)
```

```{r select-attrs}
sel <- selectFeatures(fit, fdr = 0.1)
cat("target FDR:", attr(sel, "targetFdr"),
    " estimated:", round(attr(sel, "estimatedFdr"), 4), "\n")
```

The full ranking is an ordinary data frame, so you can still inspect it
directly:

```{r summary-subset}
s <- summary(fit)
head(s[, c("Feature", "eMIP", "PostMean", "MeanNModels")], 5)
```

### `plot(fit, type = "trace")` — should I believe it?

```{r trace, fig.height = 6}
plot(fit, type = "trace")
```

You want all three panels to look like **fuzzy horizontal bands**. A visible
trend, or long flat stretches, means the chain has not settled: raise `iter` and
`burnin`.

## 4. Getting numbers out

### `coef()`

Original data scale by default, intercept included.

```{r coef}
round(coef(fit), 3)
```

### `predict()`

Predictions average over the posterior rather than plugging in one coefficient
vector. There are three levels of detail.

```{r predict}
newX <- X[1:4, ]

predict(fit, newX)                                    # point predictions
```

```{r predict-ci}
round(predict(fit, newX, interval = "credible"), 2)   # uncertainty in the mean
```

```{r predict-pi}
round(predict(fit, newX, interval = "prediction"), 2) # uncertainty in a new observation
```

Which one you want:

* **`"credible"`** — "where is the average response for this x?" Narrow.
* **`"prediction"`** — "where will the *next actual observation* at this x
  fall?" Wider, because it includes the irreducible noise.

If you are telling someone what to expect for one new case, you want
`"prediction"`.

## 5. The plots

### `type = "emip"` — which features are in?

```{r plot-emip}
plot(fit, type = "emip", top = 8)
```

### `type = "intervals"` — how big, and how certain?

```{r plot-intervals}
plot(fit, type = "intervals", top = 8)
```

Bars that cross the red zero line are unresolved. `V1` and `V2` are both clearly
non-zero but individually imprecise — the collinearity, shown honestly.

The two remaining types, `"coallocation"` and `"allocation"`, describe the
ensemble structure and appear in Section 7, where there is a properly tuned fit
to apply them to.

## 6. Choosing `lambda2` and `numModels`

Use `cv.bre()`, which scores held-out predictive error of the ensemble over a
grid and refits at the winner.

```{r cv}
cv <- cv.bre(X, y, lambda2 = c(0, 2, 5, 10, 20), numModels = 3, nfolds = 5,
             iter = 1200, burnin = 600, verbose = FALSE, seed = 3)
cv
```

Note the grid reaches much further than the default of 2. `lambda2` competes on
the log-odds scale against the evidence for including a feature, which grows
with sample size and signal strength, so the value that separates a correlated
group on one dataset can be an order of magnitude off on another. **Always
widen the grid until the chosen value is interior**, and if the largest value
wins, extend it.

`lambda2 = 0` is always worth including: it is the same model with the repulsion
switched off, so it tells you whether the repulsion is earning its keep.

```{r cv-plot, fig.height = 4}
plot(cv)
```

The refitted model is in `cv$fit` and behaves like any other fit:

```{r cv-fit}
round(coef(cv$fit)[1:4], 3)
```

### Why not WAIC?

`waic()` exists but **must not be used to choose `lambda2`**. The objective
being fitted scores each sub-model on how well it predicts *on its own*, and
repulsion necessarily makes each individual sub-model worse even when it makes
the ensemble better. WAIC therefore improves monotonically as `lambda2` goes to
zero and would always tell you to switch the method off. Held-out error of the
*ensemble* is the right target, which is what `cv.bre()` uses.

## 7. Reading the allocation structure

This is what the method offers that a single sparse model cannot.

We use the cross-validated fit from Section 6, not the under-repelled default.

`coallocation()` gives the posterior probability that two features sit in the
same sub-model. Compare it against the baseline `1/G` you would get if features
were sprinkled across sub-models at random:

```{r coalloc}
best <- cv$fit

cm <- coallocation(best, top = 4)
round(cm, 2)

cat("baseline 1/G =", round(1 / best$numModels, 2), "\n")
```

| Value | Meaning |
|---|---|
| **well below `1/G`** | the ensemble is actively separating these two. They are **substitutes** — carrying nearly the same information, so no sub-model needs both. |
| **near `1/G`** | no detectable relationship. |
| **well above `1/G`** | they are **complements**, used together; or at least one is carried by several sub-models at once. |

Read this matrix as two findings at once. `V1` and `V2` sit **below** the
baseline: they are near-duplicates, and the ensemble is using them as
substitutes in different sub-models. `V3` sits at the top of the range with
everything: it carries information nothing else does, so **every** sub-model
needs it — a complement, not a substitute.

**An essential check.** Low co-allocation has two possible causes, and only one
of them is interesting:

* both features keep a **high `eMIP`** — they really are being separated;
* one has a **low `eMIP`** — it was simply dropped. That is the right answer for
  an exact duplicate, but it is not discovered structure.

So always read co-allocation together with the inclusion probabilities, which
are on the diagonal:

```{r coalloc-check}
round(diag(cm), 3)       # these are the eMIPs
```

The exact baseline also depends on how often each feature is carried: for
features in `m_j` and `m_k` sub-models it is
`1 - choose(G - m_j, m_k) / choose(G, m_k)`, equal to `1/G` only when both are
carried exactly once.

### How the structure depends on `lambda2`

```{r lambda-sweep}
sweep <- t(sapply(c(0, 5, 10, 20), function(lam) {
    f <- bre(X, y, numModels = 3, lambda2 = lam, iter = 1200, burnin = 600,
             verbose = FALSE, seed = 7)
    c(lambda2 = lam,
      coalloc_V1_V2 = unname(coallocation(f, features = c("V1", "V2"))["V1", "V2"]),
      meanNModels_V1 = unname(rowSums(f$pip)[1]),
      eMIP_V1 = f$emip[[1]])
}))
round(sweep, 3)
```

Below a threshold the repulsion does nothing at all; above it the pair separates
cleanly while both features stay in the model. That threshold is
problem-specific, which is why Section 6 exists.

### Confirming the repulsion is doing it

Refit with the repulsion switched off and the separation should disappear:

```{r coalloc-ablation}
noRep <- bre(X, y, numModels = 3, lambda2 = 0, iter = 3000, burnin = 1500,
             verbose = FALSE, seed = 7)

round(c(withRepulsion = coallocation(best, features = c("V1", "V2"))["V1", "V2"],
        noRepulsion   = coallocation(noRep, features = c("V1", "V2"))["V1", "V2"],
        baseline      = 1 / 3), 3)
```

### One caveat: label symmetry

Permuting all `G` sub-model labels at once leaves the posterior unchanged, so
"Model 2" has no meaning across runs. `coallocation()` is unaffected by this.
`allocation()` removes it by relabelling each draw, and reports how often
relabelling was needed:

```{r allocation}
al <- allocation(best)
cat("relabelling rate:", round(al$switchRate, 3), "\n")
round(al$pip[1:4, ], 2)
```

The two structural plots, on the cross-validated fit:

```{r plot-structure, fig.height = 4}
plot(best, type = "coallocation", top = 6)
```

```{r plot-allocation, fig.height = 4}
plot(best, type = "allocation", top = 6)
```

A relabelling rate near zero is fine, and usually means the assignment is
sharply identified — the posterior really does prefer one arrangement. What you
should check instead is that the relocation move is running at all, which
`print(fit)` reports as `Transfer acc.`; if that is 0%, raise `transferMoves`.

## 8. Classification

Same interface with `family = "binomial"`. The example below is deliberately
imbalanced.

```{r binomial}
set.seed(11)
nb <- 300
Xb <- matrix(rnorm(nb * 8), nb, 8)
Xb[, 2] <- Xb[, 1] + rnorm(nb, sd = 0.15)
colnames(Xb) <- paste0("V", 1:8)

yb <- rbinom(nb, 1, plogis(-1.5 + as.numeric(Xb %*% c(1, 1, -1.5, rep(0, 5)))))
mean(yb)

fitb <- bre(Xb, yb, family = "binomial", numModels = 3,
            iter = 2500, burnin = 1250, verbose = FALSE, seed = 3)

round(coef(fitb)[1:4], 3)       # true intercept -1.5, V1 = 1, V2 = 1, V3 = -1.5
```

As before, `V1` and `V2` are the collinear pair, so read their sum:

```{r binomial-pair}
cat("V1 + V2 estimated:", round(sum(coef(fitb)[2:3]), 3), " true: 2\n")
```

For probabilities use `type = "response"`:

```{r binomial-predict}
probs <- predict(fitb, Xb, type = "response")
round(head(probs, 4), 3)

round(head(predict(fitb, Xb, type = "response", interval = "credible"), 4), 3)
```

These are posterior mean probabilities: the inverse logit is applied inside each
draw and then averaged, which is not the same as transforming the mean
coefficient. `type = "link"` gives log-odds. Prediction intervals are
Gaussian-only.

```{r binomial-calib}
cat("mean predicted:", round(mean(probs), 3),
    " observed rate:", round(mean(yb), 3), "\n")
```

## 9. The model, briefly

Each of the `G` sub-models is fitted to the response in full, and their
predictions are averaged:

```
p(y | B, sigma^2)  proportional to  prod_g  N(y | a0 + X beta^(g), sigma^2)^eta

prediction:  a0 + X %*% betaBar,     betaBar = mean of beta^(g)
```

This is a composite (Gibbs) posterior rather than an ordinary one; `eta` is the
learning rate, which sets how much evidence each sub-model draws from the data.

Each cell of the `p` by `G` allocation matrix gets a spike-and-slab prior, and
each row gets an Ising prior whose conditional inclusion odds are

```
odds(gamma_j^(g) = 1)  =  theta_j / (1 - theta_j)  *  exp(-lambda2 * c)
```

where `c` counts the *other* sub-models already using feature `j`. Each extra
sub-model holding feature `j` multiplies the odds by `exp(-lambda2)`.

### Arguments you might actually change

| Argument | Meaning | Guidance |
|---|---|---|
| `numModels` | `G`, the number of sub-models | Choose with `cv.bre()`. More sub-models can separate larger correlated groups. |
| `lambda2` | repulsion strength | Choose with `cv.bre()`. `0` is the matched control. Too large and features get pushed out of the model entirely. |
| `learningRate` | `eta` above | Default 1: every sub-model uses the data in full. Lower values temper each sub-model; `1/G` is appealing in theory but can suppress selection entirely when `p` is large. |
| `tauSq` | slab variance | On the scale of an ordinary regression coefficient, independent of `G`. Default 1 suits standardised data. |
| `iter`, `burnin`, `thin` | chain length and storage | Raise if traces look unsettled. `thin` only saves memory. |
| `transferMoves` | feature relocation proposals per iteration | Default 10. Raise if the relabelling rate or transfer acceptance is near zero. |
| `thetaShape1`, `thetaShape2` | Beta prior on inclusion | Default implies about one active feature per sub-model. Raise `thetaShape1` for denser models. |
| `pgTrunc` | Polya-Gamma series truncation | Binomial only, and only used when `learningRate` is not 1. |
| `scaleData`, `seed` | standardise `X`; reproducibility | Keep `scaleData = TRUE`; set `seed` for anything you report. |

## 10. Practical notes

**Large `p`.** Cost per iteration is `O(G * n * p)` with no matrix inversions.
Memory is the usual constraint, so use `burnin` and `thin`:

```r
fit <- bre(X, y, numModels = 5, iter = 20000, burnin = 10000, thin = 10)
```

`bre()` estimates the footprint before allocating and stops with instructions if
it exceeds `maxMemoryGb` (default 4).

**Troubleshooting.**

| Symptom | Likely cause and fix |
|---|---|
| Error about constant columns | A column of `X` has zero variance. Remove it. |
| Everything has `eMIP` near 0 | Signal too weak; or `learningRate` too low for this `p`; or `lambda2` too high. Try `lambda2 = 0` first to isolate it. |
| Everything has `eMIP` near 1 | Prior too permissive. Lower `thetaShape1`, or raise `thetaShape2`. |
| Co-allocation all near 1 | The repulsion is not biting. Raise `lambda2`, and check `cv.bre()`. |
| Transfer acceptance 0% | Features cannot relocate. Raise `transferMoves`; check `lambda2` is not enormous. |
| Traces drift | Raise `iter` and `burnin`. |
| Binomial fitting is slow | `learningRate` other than 1 needs the series Polya-Gamma sampler. Lower `pgTrunc`, or use the default rate. |
| Memory error on start | Increase `thin`, or set `storeChains = FALSE`. |

**Reproducibility.** Pass `seed` to `bre()` rather than calling `set.seed()`
beforehand, so the seed is recorded in the call.

## 11. Where to look next

* `?bre` — every argument, the model, and what the repulsion identifies.
* `?cv.bre` — choosing `lambda2`, `numModels` and `learningRate`.
* `?coallocation`, `?allocation` — allocation summaries.
* `?summary.bre`, `?predict.bre`, `?plot.bre` — method details.
